### 2019-10-11 README.Rmd
initial skeleton

---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Explore and Evaluate the Mechanisms for Missing Data in the Enhanced Tuberculosis Surveillance System

[![badge](https://img.shields.io/badge/Launch-Analysis-lightblue.svg)](https://mybinder.org/v2/gh/seabbs/ETSMissing/master?urlpath=rstudio)
[![Documentation](https://img.shields.io/badge/Documentation-click here!-lightgrey.svg?style=flat)](https://www.samabbott.co.uk/ETSMissing)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.2635687.svg)](https://doi.org/10.5281/zenodo.2635687)

[Sam Abbott](https://www.samabbott.co.uk), Hannah Christensen, Ellen Brooks-Pollock

## Objectives

## Design

## Setting

England and Wales.

## Outcome measures

## Results

## Conclusions

## Reproducibility

### Repository structure

The repository is structured as an R package. It has the following structure:

- `data-raw`: Raw data processing.
- `data`: Processed data.
- `R`: Supporting R functions.
- `docs:` Documentation for R code.
- `vignettes`: Analysis paper, results, and analysis plan.
- `peer-review`: Documentation required for peer review.

### Manual install

- Install R (analysis run with `3.6.1`) and Rstudio (alternatively use Docker as outlined below).

- Download the analysis folder from [https://github.com/seabbs/ETSMissing/archive/master.zip](https://github.com/seabbs/ETSMissing/archive/master.zip) or use `git clone`, as follows, in the command line (not the R terminal).

```{bash clone-package, eval = FALSE}
git clone https://github.com/seabbs/ETSMissing.git
```

- Once this has been downloaded click on the project file (`ETSMissing.Rproj`).

- Install the analysis dependencies and build the package using the following. To enable more robust reproducibility consider using the [`checkpoint`](https://cran.r-project.org/web/packages/checkpoint/index.html) package versioned locked to R `3.6.1`.

```{r get-packages, eval = FALSE}
#install.packages("devtools")
# To build locally
devtools::install_dev_deps(dependencies = TRUE)
devtools::install()
# Alternatively to remote install
devtools::install_github("seabbs/ETSMissing", dependencies = TRUE)
```

- Load the analysis results by running `vignettes/paper.Rmd`. Alternatively the complete analysis (along with documentation) can be reconstructed using `make` in the project root directory.

- See `data-raw` for data processing and the documentation for implementation details.

### Docker

This analysis was developed in a docker container based on the tidyverse docker image. To run the docker image run:

```{bash, eval = FALSE}
docker run -d -p 8787:8787 --name etsmissing -e USER=etsmissing -e PASSWORD=etsmissing seabbs/etsmissing
```

The rstudio client can be found on port :8787 at your local machines ip. The default username:password is assessbcgpolicychange:assessbcgpolicychange, set the user with -e USER=username, and the password with - e PASSWORD=newpasswordhere. The default is to save the analysis files into the user directory.

If you have access to the required underlying raw data (see [`tbinenglanddataclean`](https://www.samabbott.co.uk/tbinenglanddataclean/)) then the entire analysis can be reproduced from scratch by adding the following to the `docker run` command, with the data saved into `data/tb_data`. The data requirements, and structure, can be found [here](https://www.samabbott.co.uk/tbinenglanddataclean/).

```{bash, eval = FALSE}
--mount type=bind,source=$(pwd)/data/tb_data,target=/home/ETSMissing/data/tb_data
```

Alternatively the analysis environment can be accessed via [binder](https://mybinder.org/v2/gh/seabbs/AssessBCGPolicyChange/master?urlpath=rstudio).

### 2019-10-11 peer-review/cover-letter.Rmd
initial skeleton

---
title: ''
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Editor,

Please find attached our manuscript, entitled “” for consideration for publication in .

In this paper ...

We found ...

We believe our manuscript will be of interest to you and the journal’s readership because ...

We do not have any issues relating to the journal policies. All authors have approved this version of the manuscript for submission and the content of the manuscript has not been published or submitted elsewhere.

With kind regards, on behalf of all the co-authors,

Sam Abbott

### 2019-10-11 data-raw/package_set_up.R

@@ -1,0 +1,34 @@
+#install.packages("devtools")
+library(devtools)
+
+# init stuffs
+use_data_raw()
+use_cran_badge()
+use_mit_license()
+use_readme_rmd()
+use_testthat()
+use_travis()
+use_roxygen_md()
+
+
+#Suggest
+use_package("pkgdown", type = "Suggests")
+use_package("devtools", type = "Suggests")
+use_package("pomp", type = "Suggests")
+#Import
+use_package("tidyverse", type = "Imports")
+use_package("tbinenglanddataclean", type = "Imports")
+use_package("prettypublisher", type = "Imports")
+use_package("ggplot2", type = "Imports")
+use_package("purrr", type = "Imports")
+
+#Vignettes
+use_vignette("analysis_plan")
+use_vignette("paper")
+use_vignette("si")
+
+##Build site, and make pkgdown
+devtools::document()
+devtools::build_vignettes()
+devtools::build()
+pkgdown::build_site()

### 2019-10-11 tests/testthat.R

@@ -1,0 +1,4 @@
+library(testthat)
+library(ETSMissing)
+
+#test_check("ETSMissing")

### 2019-10-15 scripts/explore-variable-missingness.R
added code for analysis

@@ -1,0 +1,153 @@
+# Function to model missing data ------------------------------------------
+
+model_var_missing <- function(df, var, confounders, confounder_names) {
+
+  df <- df %>%
+    select_at(.vars = c(var, confounders)) %>%
+    mutate_at(.vars = var, funs(case_when(is.na(.) ~ "Missing",
+                                          TRUE ~ "Complete") %>%
+                                  factor(levels = c("Complete", "Missing")))) %>%
+    drop_na()
+
+  model <- glm(as.formula(paste0(var, " ~ .")), data = df, family = binomial(link="logit"))
+
+  ## Tidy model output table
+  table <- model %>%
+    tidy(conf.int = TRUE) %>%
+    mutate_at(.vars = c("estimate", "conf.low", "conf.high"),
+              exp) %>%
+    slice(-1) %>%
+    mutate(`Odds Ratio` = pretty_ci(estimate, conf.low, conf.high, sep = ", "),
+           `P value (Wald)` = signif(p.value, 3)) %>%
+    select(-estimate, -std.error, -statistic, -p.value, -conf.low, -conf.high)
+
+
+  ## Get clean variable names and demographic data
+  data <- map2_dfr(confounders, confounder_names, ~ group_by(df, .dots = c(.x, var)) %>%
+                     count %>%
+                     group_by(.dots = .x) %>%
+                     add_count(wt = n, name = "nn") %>%
+                     filter_at(.vars = var, all_vars(. == "Missing")) %>%
+                     select(-contains(var)) %>%
+                     drop_na %>%
+                     ungroup %>%
+                     rename_if(is.factor, funs(paste0("Category"))) %>%
+                     mutate_if(is.factor, as.character) %>%
+                     mutate(Variable = .y) %>%
+                     mutate(key = paste0(.x, Category)) %>%
+                     mutate(Missing = pretty_round(n / nn * 100, 1)) %>%
+                     mutate(Missing = paste0(Missing, "% (", n, ")")) %>%
+                     select(Variable, Category, `Missing (N)` = Missing,
+                            Notifications = nn, key))
+
+  lik_tests <- tibble(Variable = confounder_names,
+                      `P value (LRT)` = map_dbl(confounders, ~anova(
+                        model,
+                        update(model, paste0(". ~ . - ", .)),
+                        test = "LRT")$`Pr(>Chi)`[2]
+                      ) %>%
[…]

### 2019-10-15 scripts/nested-missing-data.R

@@ -1,0 +1,119 @@
+# Select variables of interest --------------------------------------------
+
+
+filt_ets <- ets %>%
+  select(caserepdate, year, sex, age, phec,
+         occat, ethgrp, ukborn, timesinceent,
+         symptonset, datediag, startedtreat, starttreatdate,
+         txenddate, pulmextrapulm, culture, sputsmear, anyres,
+         prevdiag, bcgvacc, bcgvaccyr, overalloutcome, tomdeathrelat,
+         dateofdeath, txenddate, natquintile)
+
+
+# Account for nested data -------------------------------------------------
+
+
+nest_filt_ets <- filt_ets %>%
+  mutate(tomdeathrelat = ifelse(overalloutcome == 'Died',
+                                tomdeathrelat %>% as.character,
+                                overalloutcome %>% as.character) %>%
+           factor) %>%
+  mutate(dateofdeath = ifelse(overalloutcome == 'Died',
+                              dateofdeath, 100000)) %>%
+  mutate(starttreatdate = ifelse(startedtreat == 'Started',
+                                 starttreatdate,
+                                 startedtreat)) %>%
+  mutate(txenddate = ifelse(overalloutcome == 'Treatment completed',
+                            txenddate,
+                            overalloutcome)) %>%
+  mutate(bcgvaccyr = ifelse(bcgvacc == 'Yes',
+                            bcgvaccyr %>% as.character,
+                            bcgvacc) %>% factor)  %>%
+  mutate(dateofdeath = ifelse(overalloutcome %in% "Died",
+                              dateofdeath,
+                              "N/A") %>%
+           as.factor) %>%
+  mutate(timesinceent = ifelse(ukborn %in% "UK born", "N/A", timesinceent))
+
+
+# Plot nested missing data ------------------------------------------------
+
+missing_strut <- nest_filt_ets %>%
+  sample_frac(0.2) %>%
+  arrange(desc(caserepdate)) %>%
+  vis_miss(warn_large_data = FALSE, sort_miss = TRUE,
+           show_perc = TRUE, cluster = FALSE) +
+  coord_flip() +
+  theme(legend.position = "bottom") +
+  theme(axis.text.x = element_text(angle = 0, hjust = 1),
+        text = element_text(size = 22))
+
[…]

### 2019-10-15 R/make_results_folder.R
adding functions

@@ -1,0 +1,32 @@
+#' Make a Results Folder
+#'
+#' @description This utility function makes a results folder with the specified name within a results folder. It will
+#' also overwrite old results if required. Returns a path to the results folder.
+#' @param prefix A character string containing the name of the results folder to create
+#' @param regen Logical - defaulting to \code{FALSE}. Should the results folders be regenerated.
+#'
+#' @return A path to the results folder.
+#' @export
+#'
+#' @importFrom here here
+#' @examples
+#'
+#' ## Code
+#' make_results_folder
+make_results_folder <- function(prefix, regen = FALSE) {
+
+  if (!dir.exists("results")) {
+    dir.create(here("vignettes", "results"))
+  }
+
+  if (!dir.exists(file.path("results", prefix)) | regen) {
+    if (regen) {
+      unlink(here(file.path("vignettes/results", prefix)), recursive = TRUE)
+    }
+    dir.create(here(file.path("vignettes", "results", prefix)))
+  }
+
+  path <- here(file.path("vignettes", "results", prefix))
+
+  return(path)
+}

### 2019-10-15 R/utils.R

@@ -2,3 +2,83 @@
 #' @export
 magrittr::`%>%`
 
+#' Save results
+#'
+#' @param df An object
+#' @param name Character string to use to save the object
+#' @param path Character string giving the folder path
+#'
+#' @return Nothing returned
+#' @export
+#'
+#' @examples
+#'
+#' ##Code
+#' save_data
+save_data <- function(df, name, path = NULL) {
+
+  path <- file.path(path, paste0(name, ".rds"))
+
+  saveRDS(df, path)
+}
+
+## Reading results
+#' Read in results
+#'
+#' @param name Character string identifying the data
+#' @inheritParams save_data
+#'
+#' @return The saved object
+#' @export
+#'
+#'
+#' @examples
+#'
+#' ##Code
+#' read_data
+read_data <- function(name, path = NULL) {
+  path <- file.path(path, paste0(name, ".rds"))
+
+  readRDS(path)
+}
+
+## Saving figures
+#' Save Figures a PNGs
+#'
+#' @param fig \code{ggplot2} object.
+#' @param name A character string giving the name for the figure
+#' @inheritParams read_data
+#' @return Nothing returned
+#' @export
+#' @importFrom ggplot2 ggsave
+#' @examples
+#'
+#' ## Code
+#' save_figure
+save_figure <- function(fig, name, path = NULL) {
+
+  path <- file.path(path, paste0(name, ".png"))
+
+  ggplot2::ggsave(path, fig, dpi = 320, width = 8, height = 8)
+}
+
+#' Load and Show Saved Figures
+#'
+#' @param name Character string identifying the figure
+#' @inheritParams save_data
+#'
+#' @return Prints a saved figure
+#' @export
+#' @inheritParams knitr include_graphics
+#'
+#' @examples
+#'
+#' ## Code
+#' show_figure
+show_figure <- function(name, path = fig_path) {
+  path <- file.path(path, paste0(name, ".png"))
+
+  plot <- knitr::include_graphics(path)
+
+  return(plot)
+}

### 2019-10-15 R/account_for_nested_missing.R
added code to funcitons:

@@ -1,0 +1,50 @@
+#' Account for Data that is Missing Due to Nesting
+#'
+#' @description This function adds entries for variables that are only missing due to nesting - rather
+#' than truely being missing. Variables that are adjusted are: cause of death, date of death, starting treatment date,
+#' treatment end date, year of BCG vaccination, and time since entry.
+#' @param df A dataframe containing cleaned ETS data as produced by \code{tbinenglanddataclean}.
+#' @return A dataframe that has nested missingness corrected for a subset of variables
+#' @export
+#' @importFrom dplyr select mutate
+#'
+#' @examples
+#'
+#'
+#' ## Code
+#' account_for_nested_missing
+account_for_nested_missing <- function(df) {
+
+  # Select variables of interest --------------------------------------------
+  filt_df <- df %>%
+    select(caserepdate, year, sex, age, phec,
+           occat, ethgrp, ukborn, timesinceent,
+           symptonset, datediag, startedtreat, starttreatdate,
+           txenddate, pulmextrapulm, culture, sputsmear, anyres,
+           prevdiag, bcgvacc, bcgvaccyr, overalloutcome, tomdeathrelat,
+           dateofdeath, txenddate, natquintile)
+
+
+  # Account for nested data -------------------------------------------------
+  nest_filt_df <- filt_df %>%
+    mutate(tomdeathrelat = ifelse(overalloutcome == 'Died',
+                                  tomdeathrelat %>% as.character,
+                                  overalloutcome %>% as.character) %>%
+             factor) %>%
+    mutate(dateofdeath = ifelse(overalloutcome == 'Died',
+                                dateofdeath, 100000)) %>%
+    mutate(starttreatdate = ifelse(startedtreat == 'Started',
+                                   starttreatdate,
+                                   startedtreat)) %>%
+    mutate(txenddate = ifelse(overalloutcome == 'Treatment completed',
+                              txenddate,
+                              overalloutcome)) %>%
+    mutate(bcgvaccyr = ifelse(bcgvacc == 'Yes',
+                              bcgvaccyr %>% as.character,
+                              bcgvacc) %>% factor)  %>%
+    mutate(dateofdeath = ifelse(overalloutcome %in% "Died",
+                                dateofdeath,
+                                "N/A") %>%
+             as.factor) %>%
+    mutate(timesinceent = ifelse(ukborn %in% "UK born", "N/A", timesinceent))
+}

### 2019-10-15 R/model_variable_missingness.R

@@ -1,0 +1,97 @@
+#' Model Variable Missingness
+#'
+#' @description
+#' @param var A character string indicating the name of the variable
+#' to explore missingness within
+#' @param confounders A character vector containing the programmatic names of variables
+#' to use as confounders.
+#' @param confounder_names A character vector containing presentation names of variables. If not supplied will default
+#' to \code{confounders}.
+#' @inheritParams account_for_nested_missing
+#' @return A dataframe containing data and model estimates for the specified variable and confounders.
+#' @export
+#' @importFrom dplyr select_at mutate_at case_when slice mutate select count add_count drop_na ungroup rename_if mutate_if group_by all_vars left_join
+#' @importFrom tidyr replace_na
+#' @importFrom stringr str_replace
+#' @importFrom prettypublisher pretty_ci pretty_round
+#' @importFrom tibble tibble
+#' @importFrom purrr map2_dfr map_dbl
+#' @importFrom broom tidy
+#' @examples
+#'
+#' ## Code
+#' model_variable_missingness
+model_variable_missingness <- function(df = NULL, var = NULL,
+                                       confounders = NULL, confounder_names = NULL) {
+
+  if (is.null(confounder_names)) {
+    confounder_names <- confounders
+  }
+
+  df <- df %>%
+    select_at(.vars = c(var, confounders)) %>%
+    mutate_at(.vars = var, funs(case_when(is.na(.) ~ "Missing",
+                                          TRUE ~ "Complete") %>%
+                                  factor(levels = c("Complete", "Missing")))) %>%
+    drop_na()
+
+  model <- glm(as.formula(paste0(var, " ~ .")), data = df, family = binomial(link="logit"))
+
+  ## Tidy model output table
+  table <- model %>%
+    tidy(conf.int = TRUE) %>%
+    mutate_at(.vars = c("estimate", "conf.low", "conf.high"),
+              exp) %>%
+    slice(-1) %>%
+    mutate(`Odds Ratio` = pretty_ci(estimate, conf.low, conf.high, sep = ", "),
+           `P value (Wald)` = signif(p.value, 3)) %>%
+    select(-estimate, -std.error, -statistic, -p.value, -conf.low, -conf.high)
+
+
+  ## Get clean variable names and demographic data
+  data <- map2_dfr(confounders, confounder_names, ~ group_by(df, .dots = c(.x, var)) %>%
+                     count %>%
+                     group_by(.dots = .x) %>%
[…]

### 2019-10-15 R/nested_missing_table.R

@@ -1,0 +1,20 @@
+#' Tabulate Nested Missing Data
+#'
+#' @description This function tabulates nested missing data using a slightly altered (for presentation) version of
+#' \code{\link[naniar](miss_var_summary)}.
+#' @inheritParams plot_nested_missing
+#' @return A dataframe summarising missing data
+#' @export
+#' @importFrom naniar miss_var_summary
+#' @importFrom dplyr mutate select
+#' @importFrom prettypublisher pretty_round
+#' @examples
+#'
+#' ## Code
+#' nested_missing_table
+nested_missing_table <- function(df) {
+  df %>%
+    miss_var_summary %>%
+    mutate(percent = pretty_round(pct_miss, digits = 1)) %>%
+    select(Variable = variable, `Missing (N)` = n_miss, `Missing (%)` = percent)
+}

### 2019-10-15 R/plot_nested_missing.R

@@ -1,0 +1,30 @@
+#' Plot Nested Missing Data
+#'
+#'
+#' @description This function plots nested missing data using a slightly altered version of
+#' \code{\link[naniar](vis_miss)}.
+#' @param df A dataframe as produced by \code{\link[ETSMissing]{account_for_nested_missing}}
+#' @param down_sample Numeric, the fraction by which to downsample the data. Defaults to 0.2.
+#'
+#' @return A plot of nested missingness
+#' @export
+#' @importFrom dplyr sample_frac arrange
+#' @importFrom naniar vis_miss
+#' @importFrom ggplot2 coord_flip theme element_text
+#' @seealso account_for_nested_missing
+#' @examples
+#'
+#' ## Code
+#' plot_nested_missing
+plot_nested_missing <- function(df, down_sample = 0.2) {
+
+  df %>%
+  sample_frac(down_sample) %>%
+    arrange(desc(caserepdate)) %>%
+    vis_miss(warn_large_data = FALSE, sort_miss = TRUE,
+             show_perc = TRUE, cluster = FALSE) +
+    coord_flip() +
+    theme(legend.position = "bottom") +
+    theme(axis.text.x = element_text(angle = 0, hjust = 1),
+          text = element_text(size = 22))
+}

### 2019-10-15 R/summarise_missingness.R

@@ -1,0 +1,58 @@
+#' Summarise Missingness by Variable
+#'
+#'
+#' @description This function filters the ETS so that only entries that could have been missing are
+#' evaluated for each variable.
+#' @inheritParams account_for_nested_missing
+#' @return A named list summarising missing data for the following variables: date of starting treatment,
+#' date of ending treatment, date of death, and cause of death.
+#' @export
+#'
+#' @importFrom dplyr filter count add_count filter mutate pull summarise
+#' @importFrom prettypublisher pretty_percentage
+#' @examples
+#'
+#' ## Code
+#' summarise_missingness
+summarise_missingness <- function(df) {
+  ## Started treatment date
+  date_treatment <- df %>%
+    filter(startedtreat %in% "Started") %>%
+    count(starttreatdate) %>%
+    add_count(wt = n, name = "nn") %>%
+    filter(is.na(starttreatdate)) %>%
+    mutate(per = pretty_percentage(n, nn, 1)) %>%
+    pull(per)
+
+  ## Ended treatment
+  date_treatment_end <- df %>%
+    filter(startedtreat %in% "Started") %>%
+    filter(overalloutcome %in% "Treatment completed") %>%
+    count(txenddate) %>%
+    add_count(wt = n, name = "nn") %>%
+    filter(is.na(txenddate)) %>%
+    mutate(per = pretty_percentage(n, nn, 1)) %>%
+    pull(per)
+
+  ## Date of death
+  date_death_missing <- df %>%
+    filter(overalloutcome == "Died") %>%
+    summarise(missing = sum(is.na(dateofdeath)), n = n()) %>%
+    mutate(per = pretty_percentage(missing, n, 1)) %>%
+    pull(per)
+
+  ## Cause of death proportion missing
+  tomdeathrelat_missing <- df %>%
+    filter(overalloutcome == "Died") %>%
+    count(tomdeathrelat) %>%
+    add_count(wt = n, name = "nn") %>%
+    mutate(pretty_per = pretty_percentage(n, nn, 1)) %>%
+    filter(is.na(tomdeathrelat)) %>%
+    pull(pretty_per)
+
+  missing_stats <- c(date_treatment, date_death_missing, tomdeathrelat_missing,
+                     date_treatment_end)
+  names(missing_stats) <- c("date_treat", "date_death", "cause_death", "date_treat_end")
+
+  return(missing_stats)
+}

### 2019-10-15 R/utils.R

@@ -20,6 +20,8 @@ save_data <- function(df, name, path = NULL) {
   path <- file.path(path, paste0(name, ".rds"))
 
   saveRDS(df, path)
+
+  invisible()
 }
 
 ## Reading results
@@ -60,6 +62,8 @@ save_figure <- function(fig, name, path = NULL) {
   path <- file.path(path, paste0(name, ".png"))
 
   ggplot2::ggsave(path, fig, dpi = 320, width = 8, height = 8)
+
+  invisible()
 }
 
 #' Load and Show Saved Figures

### 2019-10-15 scripts/explore-variable-missingness.R

@@ -1,153 +1,0 @@
-# Function to model missing data ------------------------------------------
-
-model_var_missing <- function(df, var, confounders, confounder_names) {
-
-  df <- df %>%
-    select_at(.vars = c(var, confounders)) %>%
-    mutate_at(.vars = var, funs(case_when(is.na(.) ~ "Missing",
-                                          TRUE ~ "Complete") %>%
-                                  factor(levels = c("Complete", "Missing")))) %>%
-    drop_na()
-
-  model <- glm(as.formula(paste0(var, " ~ .")), data = df, family = binomial(link="logit"))
-
-  ## Tidy model output table
-  table <- model %>%
-    tidy(conf.int = TRUE) %>%
-    mutate_at(.vars = c("estimate", "conf.low", "conf.high"),
-              exp) %>%
-    slice(-1) %>%
-    mutate(`Odds Ratio` = pretty_ci(estimate, conf.low, conf.high, sep = ", "),
-           `P value (Wald)` = signif(p.value, 3)) %>%
-    select(-estimate, -std.error, -statistic, -p.value, -conf.low, -conf.high)
-
-
-  ## Get clean variable names and demographic data
-  data <- map2_dfr(confounders, confounder_names, ~ group_by(df, .dots = c(.x, var)) %>%
-                     count %>%
-                     group_by(.dots = .x) %>%
-                     add_count(wt = n, name = "nn") %>%
-                     filter_at(.vars = var, all_vars(. == "Missing")) %>%
-                     select(-contains(var)) %>%
-                     drop_na %>%
-                     ungroup %>%
-                     rename_if(is.factor, funs(paste0("Category"))) %>%
-                     mutate_if(is.factor, as.character) %>%
-                     mutate(Variable = .y) %>%
-                     mutate(key = paste0(.x, Category)) %>%
-                     mutate(Missing = pretty_round(n / nn * 100, 1)) %>%
-                     mutate(Missing = paste0(Missing, "% (", n, ")")) %>%
-                     select(Variable, Category, `Missing (N)` = Missing,
-                            Notifications = nn, key))
-
-  lik_tests <- tibble(Variable = confounder_names,
-                      `P value (LRT)` = map_dbl(confounders, ~anova(
-                        model,
-                        update(model, paste0(". ~ . - ", .)),
-                        test = "LRT")$`Pr(>Chi)`[2]
-                      ) %>%
[…]

### 2019-10-15 scripts/nested-missing-data.R

@@ -1,119 +1,0 @@
-# Select variables of interest --------------------------------------------
-
-
-filt_ets <- ets %>%
-  select(caserepdate, year, sex, age, phec,
-         occat, ethgrp, ukborn, timesinceent,
-         symptonset, datediag, startedtreat, starttreatdate,
-         txenddate, pulmextrapulm, culture, sputsmear, anyres,
-         prevdiag, bcgvacc, bcgvaccyr, overalloutcome, tomdeathrelat,
-         dateofdeath, txenddate, natquintile)
-
-
-# Account for nested data -------------------------------------------------
-
-
-nest_filt_ets <- filt_ets %>%
-  mutate(tomdeathrelat = ifelse(overalloutcome == 'Died',
-                                tomdeathrelat %>% as.character,
-                                overalloutcome %>% as.character) %>%
-           factor) %>%
-  mutate(dateofdeath = ifelse(overalloutcome == 'Died',
-                              dateofdeath, 100000)) %>%
-  mutate(starttreatdate = ifelse(startedtreat == 'Started',
-                                 starttreatdate,
-                                 startedtreat)) %>%
-  mutate(txenddate = ifelse(overalloutcome == 'Treatment completed',
-                            txenddate,
-                            overalloutcome)) %>%
-  mutate(bcgvaccyr = ifelse(bcgvacc == 'Yes',
-                            bcgvaccyr %>% as.character,
-                            bcgvacc) %>% factor)  %>%
-  mutate(dateofdeath = ifelse(overalloutcome %in% "Died",
-                              dateofdeath,
-                              "N/A") %>%
-           as.factor) %>%
-  mutate(timesinceent = ifelse(ukborn %in% "UK born", "N/A", timesinceent))
-
-
-# Plot nested missing data ------------------------------------------------
-
-missing_strut <- nest_filt_ets %>%
-  sample_frac(0.2) %>%
-  arrange(desc(caserepdate)) %>%
-  vis_miss(warn_large_data = FALSE, sort_miss = TRUE,
-           show_perc = TRUE, cluster = FALSE) +
-  coord_flip() +
-  theme(legend.position = "bottom") +
-  theme(axis.text.x = element_text(angle = 0, hjust = 1),
-        text = element_text(size = 22))
-
[…]

### 2019-10-16 R/model_variable_missingness.R
streamlined and documented code

@@ -1,17 +1,22 @@
 #' Model Variable Missingness
 #'
-#' @description
+#' @description This function models missingness within a binary variable using logistic regression, adjusted for a specified
+#' list of confounders (only categorical variables are supported). It returns a structured output that can be directly tabulated.
+#'  P values from Wald and Likelihood tests are also returned. Odds ratios are computed and presented with their 95% confidence
+#'   intervals.
 #' @param var A character string indicating the name of the variable
-#' to explore missingness within
+#' to explore missingness within.
 #' @param confounders A character vector containing the programmatic names of variables
-#' to use as confounders.
+#' to use as confounders. All variables must be categorical.
 #' @param confounder_names A character vector containing presentation names of variables. If not supplied will default
 #' to \code{confounders}.
+#' @param conf_int_sep A character string indicating the separator to use for confidence intervals
 #' @inheritParams account_for_nested_missing
 #' @return A dataframe containing data and model estimates for the specified variable and confounders.
 #' @export
-#' @importFrom dplyr select_at mutate_at case_when slice mutate select count add_count drop_na ungroup rename_if mutate_if group_by all_vars left_join
-#' @importFrom tidyr replace_na
+#' @importFrom dplyr select_at mutate_at filter_at case_when slice mutate select count add_count ungroup rename_if mutate_if group_by all_vars left_join funs n
+#' @importFrom tidyr replace_na drop_na
+#' @importFrom tidyselect contains
 #' @importFrom stringr str_replace
 #' @importFrom prettypublisher pretty_ci pretty_round
 #' @importFrom tibble tibble
@@ -22,7 +27,8 @@
 #' ## Code
 #' model_variable_missingness
 model_variable_missingness <- function(df = NULL, var = NULL,
-                                       confounders = NULL, confounder_names = NULL) {
+                                       confounders = NULL, confounder_names = NULL,
+                                       conf_int_sep = ", ") {
 
   if (is.null(confounder_names)) {
     confounder_names <- confounders
@@ -43,7 +49,7 @@ model_variable_missingness <- function(df = NULL, var = NULL,
     mutate_at(.vars = c("estimate", "conf.low", "conf.high"),
               exp) %>%
     slice(-1) %>%
-    mutate(`Odds Ratio` = pretty_ci(estimate, conf.low, conf.high, sep = ", "),
[…]

### 2019-10-16 R/nested_missing_table.R

@@ -1,7 +1,7 @@
 #' Tabulate Nested Missing Data
 #'
 #' @description This function tabulates nested missing data using a slightly altered (for presentation) version of
-#' \code{\link[naniar](miss_var_summary)}.
+#' \code{\link[naniar]{miss_var_summary}}.
 #' @inheritParams plot_nested_missing
 #' @return A dataframe summarising missing data
 #' @export

### 2019-10-16 R/plot_nested_missing.R

@@ -2,7 +2,7 @@
 #'
 #'
 #' @description This function plots nested missing data using a slightly altered version of
-#' \code{\link[naniar](vis_miss)}.
+#' \code{\link[naniar]{vis_miss}}.
 #' @param df A dataframe as produced by \code{\link[ETSMissing]{account_for_nested_missing}}
 #' @param down_sample Numeric, the fraction by which to downsample the data. Defaults to 0.2.
 #'
@@ -19,7 +19,7 @@
 plot_nested_missing <- function(df, down_sample = 0.2) {
 
   df %>%
-  sample_frac(down_sample) %>%
+    sample_frac(down_sample) %>%
     arrange(desc(caserepdate)) %>%
     vis_miss(warn_large_data = FALSE, sort_miss = TRUE,
              show_perc = TRUE, cluster = FALSE) +

### 2019-10-16 R/utils.R

@@ -73,7 +73,7 @@ save_figure <- function(fig, name, path = NULL) {
 #'
 #' @return Prints a saved figure
 #' @export
-#' @inheritParams knitr include_graphics
+#' @importFrom knitr include_graphics
 #'
 #' @examples
 #'
@@ -85,4 +85,25 @@ show_figure <- function(name, path = fig_path) {
   plot <- knitr::include_graphics(path)
 
   return(plot)
+}
+
+#' Pull out Results by Variable
+#'
+#' @param df A dataframe containing an outcome variable (character) and a missing_table variable.
+#' @param var A character string identifying which variable to pull results for.
+#'
+#' @return A dataframe of results
+#' @export
+#'
+#' @importFrom dplyr filter pull first
+#' @examples
+#'
+#' ## Code
+#' pull_results
+pull_results <- function(df, var) {
+  df %>%
+    dplyr::filter(outcome %in% var) %>%
+    dplyr::pull(missing_table) %>%
+    dplyr::first()
+
 }

### 2019-10-16 R/account_for_nested_missing.R
dealt with package notes

@@ -15,6 +15,13 @@
 #' account_for_nested_missing
 account_for_nested_missing <- function(df) {
 
+  caserepdate <- NULL; year <- NULL; sex <- NULL; age <- NULL; phec <- NULL;
+  occat <- NULL; ethgrp <- NULL; ukborn <- NULL; timesinceent <- NULL;
+  symptonset <- NULL; datediag <- NULL; startedtreat <- NULL; starttreatdate <- NULL;
+  txenddate <- NULL; pulmextrapulm <- NULL; culture <- NULL; sputsmear <- NULL; anyres <- NULL;
+  prevdiag <- NULL; bcgvacc <- NULL; bcgvaccyr <- NULL; overalloutcome <- NULL; tomdeathrelat <- NULL;
+  dateofdeath <- NULL; txenddate <- NULL; natquintile <- NULL;
+
   # Select variables of interest --------------------------------------------
   filt_df <- df %>%
     select(caserepdate, year, sex, age, phec,

### 2019-10-16 R/model_variable_missingness.R

@@ -22,6 +22,7 @@
 #' @importFrom tibble tibble
 #' @importFrom purrr map2_dfr map_dbl
 #' @importFrom broom tidy
+#' @importFrom stats as.formula binomial glm
 #' @examples
 #'
 #' ## Code
@@ -29,6 +30,10 @@
 model_variable_missingness <- function(df = NULL, var = NULL,
                                        confounders = NULL, confounder_names = NULL,
                                        conf_int_sep = ", ") {
+
+  estimate <- NULL; conf.low <- NULL; conf.high <- NULL; p.value <- NULL;
+  std.error <- NULL; statistic <- NULL; key <- NULL; `Odds Ratio` <- NULL;
+  Variable <- NULL; dummy <- NULL; `P value (LRT)` <- NULL; `P value (Wald)` <- NULL;
 
   if (is.null(confounder_names)) {
     confounder_names <- confounders

### 2019-10-16 R/nested_missing_table.R

@@ -13,6 +13,11 @@
 #' ## Code
 #' nested_missing_table
 nested_missing_table <- function(df) {
+
+  pct_miss <- NULL; Variable <- NULL; variable <- NULL;
+  `Missing (N)` <- NULL; `Missing (%)` <- NULL; percent <- NULL;
+  n_miss <- NULL;
+
   df %>%
     miss_var_summary %>%
     mutate(percent = pretty_round(pct_miss, digits = 1)) %>%

### 2019-10-16 R/plot_nested_missing.R

@@ -8,7 +8,7 @@
 #'
 #' @return A plot of nested missingness
 #' @export
-#' @importFrom dplyr sample_frac arrange
+#' @importFrom dplyr sample_frac arrange desc
 #' @importFrom naniar vis_miss
 #' @importFrom ggplot2 coord_flip theme element_text
 #' @seealso account_for_nested_missing
@@ -17,6 +17,8 @@
 #' ## Code
 #' plot_nested_missing
 plot_nested_missing <- function(df, down_sample = 0.2) {
+
+  caserepdate <- NULL;
 
   df %>%
     sample_frac(down_sample) %>%

### 2019-10-16 R/summarise_missingness.R

@@ -15,6 +15,9 @@
 #' ## Code
 #' summarise_missingness
 summarise_missingness <- function(df) {
+  starttreatdate <- NULL; startedtreat <- NULL; nn <- NULL;
+  per <- NULL; overalloutcome <- NULL; pretty_per <- NULL;
+  dateofdeath <- NULL; tomdeathrelat <- NULL; txenddate <- NULL;
   ## Started treatment date
   date_treatment <- df %>%
     filter(startedtreat %in% "Started") %>%

### 2019-10-16 R/utils.R

@@ -79,7 +79,7 @@ save_figure <- function(fig, name, path = NULL) {
 #'
 #' ## Code
 #' show_figure
-show_figure <- function(name, path = fig_path) {
+show_figure <- function(name, path = NULL) {
   path <- file.path(path, paste0(name, ".png"))
 
   plot <- knitr::include_graphics(path)
@@ -101,6 +101,8 @@ show_figure <- function(name, path = fig_path) {
 #' ## Code
 #' pull_results
 pull_results <- function(df, var) {
+  outcome <- NULL; missing_table <- NULL;
+
   df %>%
     dplyr::filter(outcome %in% var) %>%
     dplyr::pull(missing_table) %>%

### 2019-10-16 README.Rmd
edits for paper content and tables for word
complete paper+analysis scaffold

@@
 
-## Objectives
+## Background
 
-## Design
+The Enhanced Tuberculosis Surveillance (ETS) system is a routine surveillance system - with a similar structure to other such systems - that collects data on all notified tuberculosis (TB) cases in England.
+It is routinely used to study the epidemiology of TB.
+Routine data often has a large amount of missing data which may not be fully accounted for when used in analyses.
+This study explores the evidence for associations between missingness in several key outcomes and demographic variables.
+Any such associations may introduce bias if not accounted for.
 
-## Setting
+## Methods
 
-England and Wales.
+* Introduce ETS
+* Data extraction and management
+* Structure of the ETS
+* Data completeness
+* Drivers of variable completeness (regression)
 
-## Outcome measures
+## Results *Copy from bottom*
 
-## Results
+* Missing structure
+* Drivers of variable completeness
 
 ## Conclusions
+
+* Surveillance data is likely to have a high degree of misising data. In the ETS missing for key outcomes is associated with demographic factors such as....
+* To avoid biasing analysis studies should make use of imputed data - rather than complete case analysis - and extend their imputation models to other demographic variables that may not be included in the analysis model.
+* This analysis should be repeated in other datasets - for this reason the code is available as an R package.
 

### 2019-10-16 README.Rmd
added doi and updated title

@@
 
-# Explore and Evaluate the Mechanisms for Missing Data in the Enhanced Tuberculosis Surveillance System
+# Evaluating the Mechanisms for Missing Data in the Enhanced Tuberculosis Surveillance System
 
@@
 [![Documentation](https://img.shields.io/badge/Documentation-click here!-lightgrey.svg?style=flat)](https://www.samabbott.co.uk/ETSMissing)
-[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.2635687.svg)](https://doi.org/10.5281/zenodo.2635687)
+[![DOI](https://zenodo.org/badge/214470301.svg)](https://zenodo.org/badge/latestdoi/214470301)
 

### 2019-10-17 R/plot_date_variable_missingness.R
edits for clarity + first draft in trends

@@ -1,0 +1,70 @@
+#' Plot Missingness for a Date Variable
+#'
+#' @param var A character string indicating the name of the date variable
+#' to explore missingness within.
+#' @param start_date A date indicating when to include data from (inclusive)
+#' @param end_date A date indicating when to exclude data from (exclusive)
+#' @inheritParams account_for_nested_missing
+#' @return A list of plots including: missing data by month and by day.
+#' @export
+#'
+#' @importFrom dplyr count mutate add_count filter
+#' @importFrom tidyr replace_na drop_na
+#' @importFrom lubridate floor_date year month mday
+#' @importFrom ggplot2 ggplot aes geom_violin geom_jitter scale_y_continuous theme_minimal labs
+#' @examples
+#'
+#' ## Code
+#' plot_date_variable_missingness
+plot_date_variable_missingness <- function(df = NULL, var = NULL,
+                                           start_date = NULL, end_date = NULL) {
+
+
+  ## Add default variable for standardised programming use
+  df$date <- df[[var]]
+
+  ## Count by day adding in zero counts when none detected
+  df_count <- df %>%
+    filter(date >= start_date, year <= end_date) %>%
+    drop_na(date) %>%
+    count(date, .drop = FALSE,
+          name = "notifications") %>%
+    mutate(notifications = notifications %>%
+             replace_na(0))
+
+
+  ## plot normalised incidence by month
+  month_plot <- df_count %>%
+    count(floor_date(date, "month"), wt = notifications) %>%
+    add_count(year(`floor_date(date, "month")`),
+              wt = n, name = "nn") %>%
+    mutate(n = n / nn) %>%
+    mutate(month = month(`floor_date(date, "month")`, label = TRUE)) %>%
+    ggplot(aes(x= month, y = n)) +
+    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
+    geom_jitter(alpha = 0.4) +
+    scale_y_continuous(labels = percent) +
+    theme_minimal() +
+    labs(x = "Month",
+         y = "Percentage of annual notifications")
+
+  ## Plot normalised incidencce by day
+  day_plot <- df_count %>%
+    count(floor_date(date, "day"), wt = notifications) %>%
+    add_count(floor_date(`floor_date(date, "day")`, "month"),
+              wt = n, name = "nn") %>%
[…]

### 2019-10-18 R/plot_date_variable_missingness.R
first full run through of trends work

@@ -4,25 +4,42 @@
 #' to explore missingness within.
 #' @param start_date A date indicating when to include data from (inclusive)
 #' @param end_date A date indicating when to exclude data from (exclusive)
+#' @param facet_by_year Logical defaults to \code{TRUE}. Should the plots be facetted by a year midpoint
+#' @param split_year The year to use as a splitting point when facetting. By default uses the mean year.
 #' @inheritParams account_for_nested_missing
 #' @return A list of plots including: missing data by month and by day.
 #' @export
 #'
-#' @importFrom dplyr count mutate add_count filter
+#' @importFrom dplyr count mutate add_count filter group_by
 #' @importFrom tidyr replace_na drop_na
 #' @importFrom lubridate floor_date year month mday
-#' @importFrom ggplot2 ggplot aes geom_violin geom_jitter scale_y_continuous theme_minimal labs
+#' @importFrom ggplot2 ggplot aes geom_violin geom_jitter scale_y_continuous theme_minimal labs facet_wrap
+#' @importFrom scales percent
+#' @importFrom purrr map
 #' @examples
 #'
 #' ## Code
 #' plot_date_variable_missingness
 plot_date_variable_missingness <- function(df = NULL, var = NULL,
-                                           start_date = NULL, end_date = NULL) {
+                                           start_date = NULL, end_date = NULL,
+                                           split_year = NULL,
+                                           facet_by_year = TRUE) {
 
+
+  year_facet <- NULL;
 
   ## Add default variable for standardised programming use
   df$date <- df[[var]]
 
+
+  ## Specify date filters if not given
+  if (is.null(start_date)) {
+    start_date <- min(df$date, na.rm = TRUE)
+  }
+
+  if (is.null(end_date)) {
+    end_date <- max(df$date, na.rm = TRUE)
+  }
   ## Count by day adding in zero counts when none detected
   df_count <- df %>%
     filter(date >= start_date, year <= end_date) %>%
@@ -33,37 +50,78 @@ plot_date_variable_missingness <- function(df = NULL, var = NULL,
              replace_na(0))
 
 
+  ## Add a date split variable - splitting based on them ean year
+  years_of_data <- df_count$date %>%
+    year() %>%
+    unique() %>%
+    as.numeric()
+
+
+   if (is.null(split_year)) {
+     split_year <- years_of_data %>%
[…]

### 2019-10-18 R/map_variable_names.R
rough draft for trends over time and updated primary

@@ -1,0 +1,44 @@
+#' Maps Coded Variable Names to Human Readable Names
+#'
+#' @description This function - specific to this analysis - changes coded
+#' variable names to be human readable using a predefined look-up.
+#' @param df A dataframe with a variable field.
+#'
+#' @return A datafame with human readable variable entries.
+#' @export
+#' @importFrom dplyr mutate case_when
+#' @examples
+#'
+#' ## Code
+#' map_variable_names
+map_variable_names <- function(df) {
+
+  dplyr::mutate(df, variable = case_when(
+    variable %in% "year" ~ "Year",
+    variable %in% "sex" ~ "Sex",
+    variable %in% "age" ~ "Age",
+    variable %in% "phec" ~ "Public Health England Centre",
+    variable %in% "occat" ~ "Occupation",
+    variable %in% "ethgrp" ~ "Ethnic group",
+    variable %in% "ukborn" ~ "UK birth status",
+    variable %in% "timesinceent" ~ "Time since entry",
+    variable %in% "symptonset" ~ "Date of symptom onset",
+    variable %in% "datediag" ~ "Date of diagnosis",
+    variable %in% "startedtreat" ~ "Started treatment",
+    variable %in% "starttreatdate" ~ "Date of starting treatment",
+    variable %in% "txenddate" ~ "Treatment end date",
+    variable %in% "pulmextrapulm" ~ "Pulmonary or extra-pulmonary TB",
+    variable %in% "culture" ~ "Culture",
+    variable %in% "startedtreat" ~ "Started treatment",
+    variable %in% "sputsmear" ~ "Sputum smear status",
+    variable %in% "anyres" ~ "Drug resistance",
+    variable %in% "prevdiag" ~ "Previous diagnosis",
+    variable %in% "bcgvacc" ~ "BCG status",
+    variable %in% "bcgvaccyr" ~ "Year of BCG vaccination",
+    variable %in% "overalloutcome" ~ "Overall outcome",
+    variable %in% "tomdeathrelat" ~ "Cause of death",
+    variable %in% "natquintile" ~ "Socio-economic status (quintiles)",
+    variable %in% "dateofdeath" ~ "Date of death",
+    variable %in% "caserepdate" ~ "Date of notification",
+  ))
+}

### 2019-10-18 R/nested_missing_table.R

@@ -15,11 +15,12 @@
 nested_missing_table <- function(df) {
 
   pct_miss <- NULL; Variable <- NULL; variable <- NULL;
-  `Missing (N)` <- NULL; `Missing (%)` <- NULL; percent <- NULL;
+  missing <- NULL; percent <- NULL; reported <- NULL;
   n_miss <- NULL;
 
   df %>%
     miss_var_summary %>%
-    mutate(percent = pretty_round(pct_miss, digits = 1)) %>%
-    select(Variable = variable, `Missing (N)` = n_miss, `Missing (%)` = percent)
+    dplyr::mutate(percent = pretty_round(pct_miss, digits = 1),
+                  reported = paste0(percent, " (", n_miss, ")")) %>%
+    dplyr::select(variable, percent, n_miss, reported)
 }

### 2019-10-18 R/plot_date_variable_missingness.R

@@ -2,8 +2,8 @@
 #'
 #' @param var A character string indicating the name of the date variable
 #' to explore missingness within.
-#' @param start_date A date indicating when to include data from (inclusive)
-#' @param end_date A date indicating when to exclude data from (exclusive)
+#' @param start_year Numeric, indicating which year to start including data from (inclusive)
+#' @param end_year Numeric, indicating which year to stop including data from (exclusive)
 #' @param facet_by_year Logical defaults to \code{TRUE}. Should the plots be facetted by a year midpoint
 #' @param split_year The year to use as a splitting point when facetting. By default uses the mean year.
 #' @inheritParams account_for_nested_missing
@@ -13,7 +13,7 @@
 #' @importFrom dplyr count mutate add_count filter group_by
 #' @importFrom tidyr replace_na drop_na
 #' @importFrom lubridate floor_date year month mday
-#' @importFrom ggplot2 ggplot aes geom_violin geom_jitter scale_y_continuous theme_minimal labs facet_wrap
+#' @importFrom ggplot2 ggplot aes geom_violin geom_jitter scale_y_sqrt scale_x_continuous theme_minimal labs facet_wrap
 #' @importFrom scales percent
 #' @importFrom purrr map
 #' @examples
@@ -21,28 +21,33 @@
 #' ## Code
 #' plot_date_variable_missingness
 plot_date_variable_missingness <- function(df = NULL, var = NULL,
-                                           start_date = NULL, end_date = NULL,
+                                           start_year = NULL, end_year = NULL,
                                            split_year = NULL,
                                            facet_by_year = TRUE) {
 
 
-  year_facet <- NULL;
+  year_facet <- NULL; notifications <- NULL; year_strat <- NULL;
+  nn <- NULL;
 
   ## Add default variable for standardised programming use
   df$date <- df[[var]]
 
 
   ## Specify date filters if not given
-  if (is.null(start_date)) {
-    start_date <- min(df$date, na.rm = TRUE)
+  if (is.null(start_year)) {
+    start_year <- df$date %>%
+      year %>%
+      min(na.rm = TRUE)
   }
 
-  if (is.null(end_date)) {
-    end_date <- max(df$date, na.rm = TRUE)
+  if (is.null(end_year)) {
+    end_year <- df$date %>%
+      year %>%
+      max(na.rm = TRUE)
   }
   ## Count by day adding in zero counts when none detected
   df_count <- df %>%
[…]

### 2019-10-21 R/model_variable_missingness.R
fixed addition of phec and added drug resistance

@@ -14,7 +14,7 @@
 #' @inheritParams account_for_nested_missing
 #' @return A dataframe containing data and model estimates for the specified variable and confounders.
 #' @export
-#' @importFrom dplyr select_at mutate_at filter_at case_when slice mutate select count add_count ungroup rename_if mutate_if group_by all_vars left_join funs n
+#' @importFrom dplyr select_at mutate_at filter_at case_when slice mutate select count add_count ungroup rename_if mutate_if group_by all_vars left_join n
 #' @importFrom tidyr replace_na drop_na
 #' @importFrom tidyselect contains
 #' @importFrom stringr str_replace
@@ -41,9 +41,9 @@ model_variable_missingness <- function(df = NULL, var = NULL,
 
   df <- df %>%
     select_at(.vars = c(var, confounders)) %>%
-    mutate_at(.vars = var, funs(case_when(is.na(.) ~ "Missing",
+    mutate_at(.vars = var, ~ case_when(is.na(.) ~ "Missing",
                                           TRUE ~ "Complete") %>%
-                                  factor(levels = c("Complete", "Missing")))) %>%
+                                  factor(levels = c("Complete", "Missing"))) %>%
     drop_na()
 
   model <- glm(as.formula(paste0(var, " ~ .")), data = df, family = binomial(link="logit"))
@@ -68,7 +68,7 @@ model_variable_missingness <- function(df = NULL, var = NULL,
                      select(-contains(var)) %>%
                      drop_na %>%
                      ungroup %>%
-                     rename_if(is.factor, funs(paste0("Category"))) %>%
+                     rename_if(is.factor, ~paste0("Category")) %>%
                      mutate_if(is.factor, as.character) %>%
                      mutate(Variable = .y) %>%
                      mutate(key = paste0(.x, Category)) %>%

### 2019-10-22 R/model_variable_missingness.R
wrapped up results discussion

@@ -89,7 +89,7 @@ model_variable_missingness <- function(df = NULL, var = NULL,
   output <- data %>%
     left_join(table, by = c("key" = "term")) %>%
     select(-key) %>%
-    mutate(`Odds Ratio` = `Odds Ratio` %>% replace_na("")) %>%
+    mutate(`Odds Ratio` = `Odds Ratio` %>% replace_na("\\*Reference\\*")) %>%
     mutate(dummy = Variable) %>%
     group_by(dummy) %>%
     mutate(Variable = c(Variable[1], rep("", n() - 1))) %>%

### 2019-10-22 R/model_variable_missingness.R
summary + conclusions -> discussion

@@ -89,7 +89,7 @@ model_variable_missingness <- function(df = NULL, var = NULL,
   output <- data %>%
     left_join(table, by = c("key" = "term")) %>%
     select(-key) %>%
-    mutate(`Odds Ratio` = `Odds Ratio` %>% replace_na("\\*Reference\\*")) %>%
+    mutate(`Odds Ratio` = `Odds Ratio` %>% replace_na("Reference")) %>%
     mutate(dummy = Variable) %>%
     group_by(dummy) %>%
     mutate(Variable = c(Variable[1], rep("", n() - 1))) %>%

### 2019-10-22 README.Rmd
added abstract draft

--- 2019-10-11 README.Rmd
@@
 
-## Objectives
+## Background
 
-## Design
+The Enhanced Tuberculosis Surveillance (ETS) system is a routine surveillance system that collects data on all notified tuberculosis (TB) cases in England and is routinely used in research.
+Routine data often has a large amount of missing data which may introduce bias to analyses.
+This study explores the evidence for associations between missingness in several outcomes and demographic variables.
+Any such associations may introduce bias if not accounted for.
 
-## Setting
+## Methods
 
-England and Wales.
-
-## Outcome measures
+We used data on all notifications in the ETS for England (2000-2015).
+We summarised the proportion of missing data and then used logistic regression to explore the associations between missingness and demographic variables for the following outcomes: drug resistance; BCG status and year; date of symptom onset, diagnosis, notification, starting treatment, finishing treatment and death; and cause of death.
+For all date variables, we visualised the distribution annually and by month, identifying variables at risk of bias.
 
@@
 
+All demographic variables considered were associated with data being missing for multiple outcomes.
+Missingness was not associated with all demographic variables for all outcomes.
+Associations that were present did not all have the same direction of effect.
+We found that the date of symptom onset and the date of ending treatment had a high proportion of cases occurring in January and on the 1st and 14th of each month indicating the presence of potential bias.
+
 ## Conclusions
+
+Missingness in outcomes was associated with multiple demographic variables.
+These associations could not be generalised, preventing a systematic approach to dealing with missingness.
+The identified associations may induce bias, meaning that complete case analysis should not be used for this data source - or other similar surveillance data.
+Our findings should be used motivate the use of imputation and to inform the specification of imputation models for similar surveillance data.
+
+## Keywords
+
+Tuberculosis, surveillance, missingness, imputation
 

### 2019-10-23 README.Rmd
abstract format, citations, future work
completed paper draft

@@
 
-# Evaluating the Mechanisms for Missing Data in the Enhanced Tuberculosis Surveillance System
+# Exploring Missing Data in the Enhanced Tuberculosis Surveillance System
 
@@
 Routine data often has a large amount of missing data which may introduce bias to analyses.
-This study explores the evidence for associations between missingness in several outcomes and demographic variables.
-Any such associations may introduce bias if not accounted for.
+
+## Aim
+
+Explore potential sources of bias from missing data for key outcomes in the ETS system.
 

### 2019-10-23 peer-review/cover_letter.Rmd
added draft cover letter

---
title: ''
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Editor,

Please find attached our manuscript, entitled “Exploring Missing Data in the Enhanced Tuberculosis Surveillance System” for consideration for publication in Eurosurveillance.

In this paper we explore missing data in the Enhanced Tuberculosis Surveillance (ETS) system. The ETS contains data on all notified Tuberculosis cases in England and is representative of surveillance data-sets more broadly. We assessed the impact of the introduction of the web-based ETS on missing data across a subset of variables. We then explored associations between missing data for several key outcomes and demographic variables. Finally, we identified potential biases due to issues with reporting in date variables by month and by day of the month.

We found that whilst data collection improved with the launch of the web-based ETS there was still substantial missing data. There was evidence of associations between missingness for multiple outcomes and all demographics variables considered. These associations differed in strength and direction between the outcomes considered indicating that the mechanisms driving missingness differed across outcomes. We found that several date variables showed evidence of reporting bias in January and on the first and 14th day of each month.

We believe our manuscript will be of interest to you and the journal’s readership because surveillance data is commonly used in analysis by authors published in your journal. Missing data is common in surveillance data and current common practice may not be sufficient to account for potential biases.

We do not have any issues relating to the journal policies. All authors have approved this version of the manuscript for submission and the content of the manuscript has not been published or submitted elsewhere.

With kind regards, on behalf of all the co-authors,

Sam Abbott

### 2019-11-20 README.Rmd
updated README + wordcounts

@@
 Missingness in outcomes was associated with multiple demographic variables.
-These associations could not be generalised, preventing a systematic approach to dealing with missingness.
-The identified associations may induce bias, meaning that complete case analysis should not be used for this data source - or other similar surveillance data.
-Our findings should be used motivate the use of imputation and to inform the specification of imputation models for similar surveillance data.
+These associations could not be generalised, preventing a systematic approach to dealing with missingness and indicating that domain knowledge is required.
+The identified associations may induce bias, meaning that multiple imputation may be beneficial for analyse using this data source - or other similar surveillance data.
+Our findings should be used to motivate the use of imputation and to inform the specification of imputation models that include auxiliary variables.
 
